---
title: "3. Manipulating Vectors and Matrices"
author: "James Woodcock"
output: 
    html_document:
        toc: true
        number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

There are a number of helpful functions that allow us to manipulate and transform vector object, some useful techniques and functions are demonstrated below.

# Subset

The subset function takes a vector or matrix, and returns a vector or matrix with the values of the original variable that meet the criteria you define.

```{r collapse = TRUE}
# Define a vector.
x <- c(1,2,3,4,5,6,7,8,9,10,11,12)

subset(x, x > 8) # Return values in x bigger than 8.

subset(x, x <= 2)

subset(x, x != 5) # Not equal to 5.

subset(x, x %in% c(1,2,3)) # Keep elements of x in c(1,2,3)

```

# Which

The <tt>which()</tt> function behaves in a similar way to <tt>subset()</tt>, however instead of returning the actual elements of the vector, it returns the **positions** of the elements which meet the criteria you set.
```{r}
x <- c(3,6,9,12,15,18)

which(x < 10)
```
We can see that we have obtained the positions in the original vector of elements less than 10.

# Sorting

It is also useful to be able to sort vectors by value.

```{r collapse=TRUE}
x <- runif(10)

sort(x, decreasing = TRUE)

y <- c("01 File.pdf", "1.2 File.pdf", "Very Important Document.pdf")

sort(y)
```

# Sequence Function

We can define a sequence in R easily using the following code. Sequences are useful when we want a fixed step size between two values, this can be particularly useful inside a "for" loop.

```{r collapse = TRUE}
x <- seq(from = 1, to = 10, by = 1)
x
y = seq(0, 1, 0.01)
print(y[1:10])
```

# Missing Values

There are a number of ways we can manipulate missing values inside vectors, either by deleting them, or changing their value.

```{r}
x <- c(1,6,2,6,NA,NA,2,5)

na.omit(x) # Remove the NA's from the vector.
```

We can also use the <tt>is.na()</tt> to return a logical vector which is true when the value is NA, and false when it is not. We can use this to assign values to the elements of a vector which are missing.

```{r collapse=TRUE}
is.na(x)

x[is.na(x)] <- mean(x[!is.na(x)]) # Assign missings to the mean of the vector.
x

```

Notice above when we calculate the mean of this vector we have to exclude the missing values in the calculation. This is because NA values are known as **contagious**, meaning that calculations involving NA's will return NA. For example:

```{r}
1 + 2 + 3 + 4 + NA
```



# Further Reading

More information [here.](https://datascienceplus.com/how-to-manipulate-the-vectors/)
